---
title: "ST5209/X Assignment 4"
format: pdf
editor: visual
author: "Sivakumar Nishaanth"
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
```

## 1. AR polynomial

Consider the AR(2) model $$
X_t = 4 + 0.5 X_{t-1} - 0.25 X_{t-2} + W_t.
$$ {#eq-ar2}

a.  What is the autoregressive polynomial?

    *For an AR(2) model of the form* $X_t = c + \phi_1 X_{t-1} + \phi_2 X_{t-2} + W_t,$ *the autoregressive polynomial is defined as* $\phi(z) = 1 - \phi_1 z - \phi_2 z^2$

    *From the given model, we identify* $\phi_1 = 0.5, \ \phi_2 = -0.25.$

    *Substituting into the polynomial,* $\phi(z) = 1 - 0.5z - (-0.25)z^2 = 1 - 0.5z + 0.25z^2.$

    $${\phi(z) = 1 - 0.5z + 0.25z^2}$$

b.  What are its roots?

    *To find the roots, solve* $1 - 0.5z + 0.25z^2 = 0.$

    *Multiply through by 4 to simplify:* $4 - 2z + z^2 = 0,$ *which can be written as* $z^2 - 2z + 4 = 0.$

    *Using the quadratic formula,* $z = \frac{2 \pm \sqrt{(-2)^2 - 4(1)(4)}}{2} = \frac{2 \pm \sqrt{4 - 16}}{2} = \frac{2 \pm \sqrt{-12}}{2}.$

    *Simplifying,* $z = \frac{2 \pm 2i\sqrt{3}}{2} = 1 \pm i\sqrt{3}.$

    *The modulus of each root is* $|z| = \sqrt{1^2 + (\sqrt{3})^2} = \sqrt{4} = 2.$

    $${z = 1 \pm i\sqrt{3}, \quad |z| = 2}$$

c.  Is this model causal? Why?

    *An AR model is causal if all roots of the autoregressive polynomial lie outside the unit circle, i.e. if* $|z| > 1.$

    *From part (b), both roots have modulus* $|z| = 2 > 1.$

    *Therefore, the process is causal.*

    *The model is causal since both roots lie outside the unit circle.*

d.  Given the representation $$
    X_t = \mu + \sum_{j=0}^\infty \psi_j W_{t-j},
    $$ solve for $\psi_0, \psi_1, \psi_2, \psi_3$. *Hint: Follow the strategy discussed in `linear_process_coefs.qmd`.*

    *We write the model as* $X_t = \mu + \sum_{j=0}^{\infty} \psi_j W{t-j}.$

    *Taking expectations on both sides of the AR(2) model and using* $\mathbb{E}[W_t]=0$*,* $\mu = 4 + 0.5\mu - 0.25\mu.$

    *Rearranging,* $\mu(1 - 0.5 + 0.25) = 4 \ \Rightarrow \ \mu = \frac{4}{0.75} = \frac{16}{3}.$

    $${\mu = \frac{16}{3}}$$

    *The coefficients* $\psi_j$ *satisfy* $\phi(z)\left(\sum_{j=0}^{\infty} \psi_j z^j\right) = 1.$

    *Substitute* $\phi(z) = 1 - 0.5z + 0.25z^2$*:* $(1 - 0.5z + 0.25z^2)\left(\sum_{j=0}^{\infty} \psi_j z^j\right) = 1.$

    *Expand and match coefficients of powers of* $z$*.*

    ${Constant\ term\ (z^0):}$$ \psi_0 = 1.$

    ${Coefficient\ of\ (z^1):}\psi_1 - 0.5\psi_0 = 0 \ \Rightarrow \ \psi_1 = 0.5.$

    ${Coefficient\ of\ (z^2):}\psi_2 -0.5\psi_1 +0.25\psi_0 = 0$

    *Substitute:* $\psi_2 - 0.5(0.5) + 0.25(1) = 0 \ \Rightarrow \ \psi_2 - 0.25 + 0.25 = 0 \ \Rightarrow \ \psi_2 = 0.$

    ${Coefficient\ of\ (z^3):} \psi_3 - 0.5\psi_2 + 0.25\psi_1 = 0.$

    *Substitute:* $\psi_3 - 0 + 0.25(0.5) = 0 \ \Rightarrow \ \psi_3 + 0.125 = 0 \ \Rightarrow \ \psi_3 = -0.125.$

    $${
    \begin{aligned}
    \psi_0 &= 1, \\
    \psi_1 &= 0.5, \\
    \psi_2 &= 0, \\
    \psi_3 &= -0.125.
    \end{aligned}
    }$$

    $$X_t = \frac{16}{3} + W_t + 0.5W_{t-1} + 0 \cdot W_{t-2} - 0.125W_{t-3} + \cdots.$$

## 2. Likelihood

Consider the AR(1) model $$
X_t = 1-0.6X_{t-1} + W_t,
$$ where $W_t \sim WN(0, 0.25)$ is Gaussian white noise. We are given observations $x_1 = 0.2, x_2 = -0.3, x_3 = 0.4$.

a.  What is the mean of $X_t$?

    *For an AR(1) model* $X_t = c + \phi X_{t-1} + W_t,$ *the mean is given by* $\mu = \frac{c}{1 - \phi}.$

    *From the model,* $c = 1, \ \phi = -0.6.$

    *Thus,* $\mu = \frac{1}{1 - (-0.6)} = \frac{1}{1.6} = \frac{5}{8}.$

    $${\mu = \frac{5}{8}}$$

b.  Write the conditional likelihood of this model, when conditioning on the value of $x_1$.

    *We condition on* $(x_1)$*, so the likelihood is based on* $X_2 \mid X_1 \ \text{and} \ X_3 \mid X_2.$

    *From the AR(1) model,* $X_t \mid X_{t-1} \sim \mathcal{N}(1 - 0.6X_{t-1},, 0.25).$

    *Hence,* $X_2 \mid X_1 = x_1 \sim \mathcal{N}(1 - 0.6x_1,, 0.25), ] [ X_3 \mid X_2 = x_2 \sim \mathcal{N}(1 - 0.6x_2,, 0.25).$

    *Therefore, the conditional likelihood is* $L = f(x_2 \mid x_1) \cdot f(x_3 \mid x_2).$

    *Using the normal density,* $f(x_t \mid x_{t-1}) = \frac{1}{\sqrt{2\pi(0.25)}} \exp\left( -\frac{(x_t - (1 - 0.6x_{t-1}))^2}{2(0.25)} \right).$

    *Thus,* $$ L = \prod_{t=2}^{3} \frac{1}{\sqrt{2\pi(0.25)}} \exp\left( -\frac{(x_t - (1 - 0.6x_{t-1}))^2}{2(0.25)} \right).$$

    *Computing the conditional means:*

    $1 - 0.6x_1 = 1 - 0.6(0.2) = 1 - 0.12 = 0.88,$

    $1 - 0.6x_2 = 1 - 0.6(-0.3) = 1 + 0.18 = 1.18.$

    *Hence,* $$ L = \frac{1}{\sqrt{2\pi(0.25)}} \exp\left( -\frac{(-0.3 - 0.88)^2}{2(0.25)} \right) \times \frac{1}{\sqrt{2\pi(0.25)}} \exp\left( -\frac{(0.4 - 1.18)^2}{2(0.25)} \right). $$

    *Simplifying:*

    $(-0.3 - 0.88)^2 = (-1.18)^2 = 1.3924, ] [ (0.4 - 1.18)^2 = (-0.78)^2 = 0.6084.$

    *So,* $L = \frac{1}{2\pi(0.25)} \exp\left( -\frac{1.3924 + 0.6084}{2(0.25)} \right).$

    $= \frac{1}{0.5\pi} \exp\left( -\frac{2.0008}{0.5} \right) = \frac{1}{0.5\pi} \exp(-4.0016).$

    $$L = \frac{1}{0.5\pi} \exp\left(-4.0016\right)$$

## 3. Reversibility

a.  Using `set.seed(5209)` to fix a random seed, create a sample trajectory of length $n = 500$ from the AR(2) model from Problem 1 using `arima.sim()`.

    *We simulate a sample path from the AR(2) model* $X_t = 4 + 0.5X_{t-1} - 0.25X_{t-2} + W_t.$

    ```{r}
    set.seed(5209)

    n <- 500
    phi1 <- 0.5
    phi2 <- -0.25
    mu <- 4 / (1 - phi1 - phi2)

    x <- arima.sim(model = list(ar = c(phi1, phi2)), n = n) + mu
    ```

b.  Reverse the time index of the vector you obtain using `rev()`.

    ```{r}
    x_rev <- rev(x)
    ```

c.  Fit an AR(2) model to the reversed time series using `fable`. Hint: You may use the code snippet `model(AR(X ~ order(2)))`.

    ```{r}
    rev_data <- tibble(
      t = 1:n,
      X = as.numeric(x_rev)
    ) |>
      as_tsibble(index = t)

    fit_rev <- rev_data |>
      model(AR(X ~ order(2)))
    ```

d.  Inspect the model parameters using `tidy()`. Why are they similar to the those in @eq-ar2?

    ```{r}
    tidy(fit_rev)
    ```

    *We extract the estimated coefficients using* $tidy()$*.*

    *From the output, the estimated parameters are:* $\phi_1 \approx 0.499, \ \phi_2 \approx -0.239,$ *and the intercept is approximately* $4.008$*.*

    *These values are very close to the true parameters of the original AR(2) model:* $\phi_1 = 0.5, \ \phi_2 = -0.25, \ c = 4.$

    *This occurs because the AR(2) model is stationary and Gaussian. For a Gaussian time series, the joint distribution is fully determined by its mean and autocovariance function. Reversing the time index does not change the autocovariance structure, so the reversed series has the same dependence structure as the original process.*

    *Therefore, fitting an AR(2) model to the reversed series yields parameter estimates that are very similar to the original model.*

e.  Make a forecast with $h = 10$. What does this correspond to in terms of the original time series?

    ```{r}
    fc_rev <- forecast(fit_rev, h = 10)
    fc_rev
    ```

    *We produce a 10-step-ahead forecast from the fitted AR(2) model for the reversed series.*

    *Since the model is fitted to the time-reversed series, forecasting* $(10)$ *steps ahead in the reversed time series corresponds to predicting* $(10)$ *steps backward in the original time series.*

    Thus, this produces $10\ backcasts$, corresponding to estimates of $X_0, X_{-1}, \dots, X_{-9}.$

    *Therefore, this forecast corresponds to a backcast* *for the original time series.*

## 4. Yule-Walker

a.  Write the Yule-Walker equations for the AR(2) model from Problem 1.

    *For an AR(2) model* $X_t = \phi_1 X_{t-1} + \phi_2 X_{t-2} + W_t,$ *the Yule--Walker equations are given by* $\gamma(k) = \phi_1 \gamma(k-1) + \phi_2 \gamma(k-2), \quad k \geq 1,$ *where* $(\gamma(k))$ *is the autocovariance function.*

    *For* $(k = 1): \gamma(1) = \phi_1 \gamma(0) + \phi_2 \gamma(1).$

    *For* $(k = 2): \gamma(2) = \phi_1 \gamma(1) + \phi_2 \gamma(0).$

    *For* $(k = 0): \gamma(0) = \phi_1 \gamma(1) + \phi_2 \gamma(2) + \sigma^2$*.*

    *For the given AR(2) model,* $\phi_1 = 0.5, \ \phi_2 = -0.25,$

    *Substituting these values, the Yule--Walker equations become: \
    *$$\gamma(1) = 0.5\gamma(0) - 0.25\gamma(1), \newline \gamma(2) = 0.5\gamma(1) - 0.25\gamma(0),\newline  \gamma(0) = 0.5\gamma(1) - 0.25\gamma(2) + \sigma^2.$$

b.  Arrange the equations in the following matrix form (i.e. fill in the missing entries): $$
    \left[\begin{matrix}     ? & ? & ?  \\ ? & ? & ? \\
    ? & ? & ?
    \end{matrix}\right] \left[\begin{matrix}     \gamma(0) \\ \gamma(1) \\ \gamma(2)
    \end{matrix}\right] = \left[\begin{matrix}     \sigma^2 \\ 0 \\ 0
    \end{matrix}\right].
    $$

    *From part (a), the Yule--Walker equations are* $\gamma(0) = 0.5\gamma(1) - 0.25\gamma(2) + \sigma^2,\ \gamma(1) = 0.5\gamma(0) - 0.25\gamma(1), \ \gamma(2) = 0.5\gamma(1) - 0.25\gamma(0).$

    *Rearranging all terms to the left-hand side gives* $\gamma(0) - 0.5\gamma(1) + 0.25\gamma(2) = \sigma^2, \ -0.5\gamma(0) + 1.25\gamma(1) + 0\gamma(2) = 0, \ 0.25\gamma(0) - 0.5\gamma(1) + \gamma(2) = 0.$

    *Hence, the matrix form is \
    *$$
    \begin{bmatrix}
    1 & -0.5 & 0.25 \\
    -0.5 & 1.25 & 0 \\
    0.25 & -0.5 & 1
    \end{bmatrix}
    \begin{bmatrix}
    \gamma(0) \\
    \gamma(1) \\
    \gamma(2)
    \end{bmatrix}
    =
    \begin{bmatrix}
    \sigma^2 \\
    0 \\
    0
    \end{bmatrix}.
    $$

c.  Solve the system from part b) for $\gamma(0), \gamma(1), \gamma(2)$ numerically using `solve()`.

    ```{r}
    A <- matrix(c(
      1,   -0.5,  0.25,
     -0.5,  1.25, 0,
      0.25,-0.5,  1
    ), nrow = 3, byrow = TRUE)

    b <- c(1, 0, 0) 

    solve(A, b)
    ```

d.  Given $$
    \Gamma_2 = \left[\begin{matrix}     \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0)
    \end{matrix}\right],
    $$ what is the top left entry of $\Gamma_2^{-1}$ in terms of $\gamma(0)$ and $\gamma(1)$?

    *We are given* $\Gamma_2 =
    \begin{bmatrix}
    \gamma(0) & \gamma(1) \\
    \gamma(1) & \gamma(0)
    \end{bmatrix}.$

    *For a* $2 \times 2$ *matrix* $\begin{bmatrix}
    a & b \\
    b & a
    \end{bmatrix},$ *the inverse is given by* $\frac{1}{a^2 - b^2}
    \begin{bmatrix}
    a & -b \\
    -b & a
    \end{bmatrix}.$

    *Applying this with* $a = \gamma(0)$ *and* $b = \gamma(1)$*, we obtain* $\Gamma_2^{-1}
    =
    \frac{1}{\gamma(0)^2 - \gamma(1)^2}
    \begin{bmatrix}
    \gamma(0) & -\gamma(1) \\
    -\gamma(1) & \gamma(0)
    \end{bmatrix}.$

    *Hence, the top-left entry of* $\Gamma_2^{-1}$ *is* $$
    {
    \frac{\gamma(0)}{\gamma(0)^2 - \gamma(1)^2}
    }.
    $$\

    ```{r}
    gamma0 <- 1.26984127
    gamma1 <- 0.50793651

    gamma0 / (gamma0^2 - gamma1^2)
    ```

e.  Write a 95% confidence interval for $\phi_1$ using your answers for d) and 3d).

    ```{r}
    phi1_hat <- 0.498733
    gamma_inv_11 <- 0.9375
    n <- 500

    se <- sqrt(gamma_inv_11 / n)
    ci <- phi1_hat + c(-1, 1) * 1.96 * se

    ci
    ```

    *From part 3(d), the estimated coefficient is* $\hat{\phi}_1 \approx 0.498733.$

    *From part 4(d), the top-left entry of* $\Gamma_2^{-1}$ *is*

    $(\Gamma_2^{-1})_{11} = 0.9375.$

    *Using the asymptotic result*

    $\operatorname{Var}(\hat{\phi}_1) \approx \frac{1}{n}(\Gamma_2^{-1})_{11},$ *with* $n=500$*, we get*

    $\operatorname{se}(\hat{\phi}_1)
    = \sqrt{\frac{0.9375}{500}}
    \approx 0.0433.$

    *Hence, a 95% confidence interval for* $\phi_1$ *is* $\hat{\phi}_1 \pm 1.96 \operatorname{se}(\hat{\phi}_1)
    = 0.498733 \pm 1.96(0.0433).$

    *Therefore,* $0.498733 \pm 0.08487$*, so the 95% confidence interval is* $${(0.4139,\;0.5836)}.$$

## 5. Recursive forecasting for ARMA(1, 1)

Consider the $ARMA(1,1)$ model $$
X_t - 0.5X_{t-1} = W_t + 0.5W_{t-1}.
$$ In this question, we will investigate recursive forecasting. The following code snippet generates a sequence of length $n=50$ drawn from the above model.

```{r}
set.seed(5209)
n <- 50
wn <- rnorm(n)
xt <- arima.sim(model = list(ar = 0.5, ma = 0.5), innov = wn, n = n)
```

a.  Fill in the following code snippet using equation (12.14) to generate a sequence `wn_hat`.

```{r}
wn_hat <- rep(0, n)
wn_hat[[1]] <- xt[[1]]
for (i in 2:n) {
  wn_hat[[i]] <- xt[[i]] - 0.5 * xt[[i - 1]] - 0.5 * wn_hat[[i - 1]]
}
```

b.  Make a time plot of the log absolute difference between `wn` and `wn_hat`.

    ```{r}
    library(forecast)

    log_diff <- log(abs(wn - wn_hat))
    autoplot(ts(log_diff)) +
      labs(
        title = "Log Absolute Difference between wn and wn_hat",
        x = "Time",
        y = "log |wn - wn_hat|"
      )
    ```

c.  What consequence does this have for truncated forecasts?

    *From part (b), the log absolute difference* $\log |W_t - \hat{W}_t|$ *decreases over time, indicating that* $\hat{W}_t \to W_t$ *as* $t$ *increases.*

    *This implies that the recursively computed innovations provide increasingly accurate approximations to the true innovations.*

    *Hence, truncating the recursion (i.e., using a finite number of past terms) has negligible impact for large* $t$*. Therefore, truncated forecasts are good approximations to the true forecasts, as the error introduced by truncation becomes negligible.*

d.  Compute the truncated forecast for $X_{53}$.

    *The ARMA(1,1) model is* $X_t = 0.5X_{t-1} + W_t + 0.5W_{t-1}.$

    *Using information up to time (t=50), the one-step-ahead forecast is* $\hat X_{51|50} = 0.5X_{50} + 0.5\hat W_{50},$

    *since* $\mathbb{E}(W_{51}\mid \mathcal{F}_{50})=0$*.*

    *For the two-step-ahead forecast,* $\hat X_{52|50} = 0.5\hat X_{51|50},$

    *because both* $W_{52}$ *and* $W_{51}$ *have conditional expectation* $0$*.*

    *Similarly, the three-step-ahead forecast is* $\hat X_{53|50} = 0.5\hat X_{52|50}.$

    *Substituting recursively,* $\hat X_{53|50}
    = 0.5^2 \hat X_{51|50}
    = 0.5^2 \bigl(0.5X_{50} + 0.5\hat W_{50}\bigr)
    = 0.125X_{50} + 0.125\hat W_{50}.$

    *Hence, the truncated forecast for* $X_{53}$ *is* $$\hat X_{53|50} = 0.125\bigl(X_{50} + \hat W_{50}\bigr)$$

    ```{r}
    x51_hat <- 0.5 * xt[50] + 0.5 * wn_hat[50]
    x52_hat <- 0.5 * x51_hat
    x53_hat <- 0.5 * x52_hat

    x53_hat
    ```

## 6. ACF, PACF, and BLPs

Let $(X_t)$ be a mean zero stationary process with the following autocovariance values:

$$
\gamma_X(0) = 2, \gamma_X(1) = 1.4, \gamma_X(2) = 0.6, \gamma_X(3) = 0.4, \gamma_X(4) = 0.2.
$$

a.  Can $(X_t)$ be an MA(2) process? Explain why or why not.

    *For an MA(2) process, the autocovariance function satisfies* $\gamma_X(h) = 0 \ \text{for all } |h| > 2.$

    *However, the given autocovariances are* $\gamma_X(3) = 0.4, \ \gamma_X(4) = 0.2,$ *which are both nonzero.*

    *Therefore,* $(X_t)$ *cannot be an MA(2) process.*

    $$\text{No, } (X_t) \text{ is not an MA(2) process.}$$

b.  Can $(X_t)$ be an AR(1) process? Explain why or why not.

    *For an AR(1) process, the autocovariance function satisfies* $\gamma(h) = \phi^h \gamma(0), \ h \geq 0,$ *so that* $\phi = \frac{\gamma(1)}{\gamma(0)}.$

    *From the given values,* $\phi = \frac{1.4}{2} = 0.7.$

    *For an AR(1) process, we must have* $\gamma(2) = \phi^2 \gamma(0) = (0.7)^2 \cdot 2 = 0.98.$

    *However, the given value is* $\gamma(2) = 0.6 \neq 0.98.$

    *Therefore, the autocovariances do not follow the AR(1) structure.*

    $$\text{No, } (X_t) \text{ is not an AR(1) process.}$$

c.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_3 = 2$?

    *Since* $(X_t)$ *is a mean zero stationary process, the best linear predictor of* $X_4$ *given* $X_3$ *is* $\hat{X}_4 = \frac{\gamma(1)}{\gamma(0)} X_3.$

    *From the given autocovariances,* $\frac{\gamma(1)}{\gamma(0)} = \frac{1.4}{2} = 0.7.$

    *Thus,* $\hat{X}_4 = 0.7 \times 2 = 1.4.$

    $$\hat{X}_4 = 1.4$$

d.  Using the notation in part c), what is the variance of $X_4 - \hat X_4$?

    *From part (c), the best linear predictor is* $\hat{X}_4 = \frac{\gamma(1)}{\gamma(0)} X_3.$

    *The variance of the prediction error is given by* $\operatorname{Var}(X_4 - \hat{X}_4)
    = \gamma(0) - \frac{\gamma(1)^2}{\gamma(0)}.$

    *Substituting the given values,* $\operatorname{Var}(X_4 - \hat{X}_4)
    = 2 - \frac{(1.4)^2}{2}
    = 2 - \frac{1.96}{2}
    = 2 - 0.98
    = 1.02.$

    $$\operatorname{Var}(X_4 - \hat{X}_4) = 1.02$$

e.  What is the best linear predictor $\hat X_4$ for $X_4$ given only $X_2 = 2$?

    *The best linear predictor of* $(X_4)$ *given* $(X_2)$ *is* $\hat{X}_4 = \frac{\gamma(2)}{\gamma(0)} X_2.$

    *From the given autocovariances,* $\frac{\gamma(2)}{\gamma(0)} = \frac{0.6}{2} = 0.3.$

    *Thus,* $\hat{X}_4 = 0.3 \times 2 = 0.6.$

    $$\hat{X}_4 = 0.6$$

f.  Using the notation in part e), what is the variance of $X_4 - \hat X_4$?

    *From part (e), the best linear predictor is* $\hat{X}_4 = \frac{\gamma(2)}{\gamma(0)} X_2.$

    *The variance of the prediction error is given by* $\operatorname{Var}(X_4 - \hat{X}_4)
    = \gamma(0) - \frac{\gamma(2)^2}{\gamma(0)}.$

    *Substituting the given values,* $\operatorname{Var}(X_4 - \hat{X}_4)
    = 2 - \frac{(0.6)^2}{2}
    = 2 - \frac{0.36}{2}
    = 2 - 0.18
    = 1.82.$

    $$\operatorname{Var}(X_4 - \hat{X}_4) = 1.82$$

g.  Let $\alpha_X$ denote the partial autocorrelation function of $(X_t)$. What is $\alpha_X(1)$?

    *The partial autocorrelation at lag 1 is given by* $\alpha_X(1) = \frac{\gamma(1)}{\gamma(0)}.$

    *From the given autocovariances,* $\alpha_X(1) = \frac{1.4}{2} = 0.7.$

    $$\alpha_X(1) = 0.7$$

h.  What is $\alpha_X(3)$?

    *The partial autocorrelation at lag 3 is the last coefficient in the best linear prediction of* $X_4$ *from* $X_3, X_2, X_1$*. Equivalently, it is the Durbin--Levinson coefficient* $\phi_{33}$*.*

    *First, write the autocorrelations:* $\rho(1)=\frac{\gamma(1)}{\gamma(0)}=\frac{1.4}{2}=0.7,\
    \rho(2)=\frac{\gamma(2)}{\gamma(0)}=\frac{0.6}{2}=0.3,\
    \rho(3)=\frac{\gamma(3)}{\gamma(0)}=\frac{0.4}{2}=0.2.$

    *From part (g),* $\phi_{11}=\alpha_X(1)=0.7.$

    *Next,* $\phi_{22}
    =
    \frac{\rho(2)-\phi_{11}\rho(1)}{1-\phi_{11}\rho(1)}
    =
    \frac{0.3-(0.7)(0.7)}{1-(0.7)(0.7)}
    =
    \frac{0.3-0.49}{1-0.49}
    =
    \frac{-0.19}{0.51}
    \approx -0.3725.$

    *Also,* $\phi_{21}
    =
    \phi_{11}-\phi_{22}\phi_{11}
    =
    0.7-(-0.3725)(0.7)
    \approx 0.9608.$

    *Therefore,* $\alpha_X(3)=\phi_{33}
    =
    \frac{\rho(3)-\phi_{21}\rho(2)-\phi_{22}\rho(1)}
    {1-\phi_{21}\rho(1)-\phi_{22}\rho(2)}.$

    *Substituting the values,* $\alpha_X(3)
    =
    \frac{0.2-(0.9608)(0.3)-(-0.3725)(0.7)}
    {1-(0.9608)(0.7)-(-0.3725)(0.3)}
    \approx 0.3929.$

    *Hence,* $$\alpha_X(3)\approx 0.3929.$$

**AI Declaration:**

ChatGPT was used as a supplementary tool to assist with conceptual clarification and LaTeX formatting during the preparation of this assignment. All mathematical derivations, computations, and final answers were carried out by the author.
