---
title: "ST5209/X Assignment 5"
format: pdf
editor: visual
author: "Sivakumar Nishaanth"
---

## Set up

1.  Make sure you have the following installed on your system: $\text{\LaTeX}$, R4.2.2+, RStudio 2023.12+, and Quarto 1.3.450+.
2.  Pull changes from the course [repo](https://github.com/yanshuotan/st5209-2024).

## Submission

1.  Render the document to get a .pdf printout.
2.  Submit both the .qmd and .pdf files to Canvas.

```{r}
#| message: FALSE
#| echo: FALSE
library(tidyverse)
library(fpp3)
```

## 1. Modeling with ARIMA (Q9.7 in Hyndman & Athanasopoulos)

Consider `aus_airpassengers`, the total number of passengers (in millions) from Australian air carriers for the period 1970-2016.

```{r}
aus_airpassengers |> 
  autoplot()
```

a.  Use `ARIMA()` to find an appropriate ARIMA model. What model was selected. Check that the residuals look like white noise. Plot forecasts for the next 10 periods.

    ```{r}
    arima_fit <- aus_airpassengers |>
      model(ARIMA(Passengers))

    report(arima_fit)
    ```

    ```{r}
    arima_fit |> gg_tsresiduals()

    augment(arima_fit) |>
      features(.innov, ljung_box, lag = 10)
    ```

    ```{r}
    fc <- arima_fit |> forecast(h = 10)

    fc |>
      autoplot(aus_airpassengers)
    ```

    *The automatic ARIMA() function selected an ARIMA(0,2,1) model for the series. Residual diagnostics, including the ACF plot and Ljung-Box test, confirm the residuals resemble white noise, indicating a good fit, and the 10-period forecast shows a continuation of the upward trend with widening prediction intervals.*

b.  Write the model in terms of the backshift operator.

    *The* $ARIMA(0,2,1)$ *model is written as:* $[ (1 - B)^2 y_t = (1 - \theta_1 B)\varepsilon_t ]$\
    Using $\theta_1 = -0.8963$, $$
    (1-B)^2 y_t = (1-0.8963B)\varepsilon_t.
    $$*Where* $(1-B)^2$ *applies second-order differencing and* $(1-\theta_1 B)$ *is the* $MA(1)$ *term.*

c.  Plot forecasts from an ARIMA(0,1,0) model with drift and compare these to part a.

    ```{r}
    fit_c <- aus_airpassengers |>
      model(ARIMA(Passengers ~ 1 + pdq(0,1,0)))

    fit_c |> forecast(h=10) |>
      autoplot(aus_airpassengers)
    ```

    $ARIMA(0,1,0)$ *with drift is a random walk with a constant upward drift; forecasts are linear but less flexible than part (a), producing a straight-line projection.*

d.  Plot forecasts from an ARIMA(2,1,2) model with drift and compare these to part a. Remove the constant and see what happens.

    ```{r}
    fit_d <- aus_airpassengers |>
      model(
        with_const    = ARIMA(Passengers ~ 1 + pdq(2,1,2)),
        without_const = ARIMA(Passengers ~ 0 + pdq(2,1,2))
      )

    fit_d |> forecast(h=10) |>
      autoplot(aus_airpassengers)
    ```

    $ARIMA(2,1,2)$ *with drift produces forecasts similar to the auto-selected* $ARIMA(0,2,1)$ *from part (a). Removing the constant causes the model to fail with a non-stationary AR error, highlighting that the constant is necessary to stabilise the* $ARIMA(2,1,2)$ *fit.*

e.  Plot forecasts from an ARIMA(0,2,1) model with a constant. What happens?

    ```{r}
    fit_e <- aus_airpassengers |>
      model(ARIMA(Passengers ~ 1 + pdq(0,2,1)))

    fit_e |> forecast(h=10) |>
      autoplot(aus_airpassengers)
    ```

    *Adding a constant to an* $ARIMA(0,2,1)$$(i.e., d=2)$ *introduces a quadratic trend into the forecasts — the predictions curve upward increasingly steeply, which is usually unrealistic for long horizons.*

## 2. Seasonal ARIMA

a.  Load `diabetes.rds` from the directory `_data/cleaned`.

    ```{r}
    diabetes <- readRDS("../_data/cleaned/diabetes.rds")
    diabetes
    ```

b.  Perform the following transformation of the column `Cost`: Apply a log transform followed by a seasonal difference. Label the resulting time series `Y`.

    ```{r}
    diabetes <- diabetes |>
      mutate(log_cost = log(Cost))

    Y <- diabetes |>
      mutate(Y = difference(log(Cost), lag = 12)) |>
      filter(!is.na(Y))

    Y 

    Y |> autoplot(Y)
    ```

c.  Apply the KPSS test to `Y`. What is its p-value? What can you conclude about `Y`?

    ```{r}
    Y |> features(Y, unitroot_kpss)
    ```

    *The KPSS test has* $H_0$ *of stationarity, so a p-value of 0.1 means we fail to reject* $H_0$*— Y is stationary. After applying the log and seasonal difference, the series has no remaining trend or seasonality.*

d.  Make a time plot of `log(Cost)`. Why does the trend disappear when we consider `Y`?

    ```{r}
    diabetes |>
      autoplot(log_cost)
    ```

    $log(Cost)$ *still exhibits an upward trend and seasonal pattern. Taking the seasonal difference removes both by subtracting each value from the same month one year prior, leaving Y stationary.*

e.  Fit an ARIMA model to `log(Cost)` and report the order of the fitted model.

    ```{r}
    fit <- diabetes |>
      model(ARIMA(log(Cost)))

    report(fit)
    ```

f.  How many fitted parameters are there in the model?

    *The fitted model is ARIMA(2,0,2)(1,1,1)\[12\] with drift, which has 2 AR coefficients (ar1, ar2), 2 MA coefficients (ma1, ma2),1 seasonal MA coefficient (sma1), 1 seasonal AR coefficient (sar1) and 1 drift constant. This gives a total of 7 fitted parameters.*

## 3. Model selection

a.  What are the null hypothesis and assumptions of the ADF test?

    *The ADF test has null hypothesis that the series contains a unit root, so it is non-stationary. The alternative is stationarity. \
    \
    It assumes the series can be represented by an autoregressive model with enough lagged differences included to remove serial correlation in the errors, and that the deterministic part of the model has been specified correctly.*

b.  Is it possible for both the ADF and KPSS test applied to a dataset to have large p-values? Explain why or why not.

    *Yes. This is possible because the ADF and KPSS tests have different null hypotheses. ADF tests non-stationarity as the null, while KPSS tests stationarity as the null. If both p-values are large, the evidence is inconclusive: neither non-stationarity nor stationarity is strongly rejected - the data is simply not informative enough to distinguish between stationary and non-stationary behaviour, i.e., the series lies in an ambiguous borderline region.*

c.  What are the AIC and AICc penalties for the model fitted in Q2?

    For the fitted model $(\text{ARIMA}(2,0,2)(1,1,1)[12])$ with drift, *which has 2 AR coefficients (ar1, ar2), 2 MA coefficients (ma1, ma2),1 seasonal MA coefficient (sma1), 1 seasonal AR coefficient (sar1) and 1 drift constant. This gives a total of 7 fitted parameters.*

    So the number of fitted parameters is:$[ k = 7 ]$

    The AIC penalty is:$[ 2k = 2(7)=14 ]$

    The AICc penalty adds the small-sample correction: $[ \frac{2k(k+1)}{n-k-1} ]$

    Using $(n=204)$: $[ \frac{2(7)(8)}{204-7-1} = \frac{112}{196} =0.571 ]$

    Therefore: $AIC \ penalty = 14,\
     Additional \ AICc \ correction = 0.571, \
     Total \ AICc \ penalty = 14.571$

d.  Fit an exponential smoothing model of your choice to `diabetes.rds`. Use `glance()` to view the log likelihood and AICc values of both this model and the ARIMA model from Q2.

    ```{r}
    fits <- diabetes |>
      model(
        arima = ARIMA(log(Cost)),
        ets   = ETS(log(Cost))
      )

    glance(fits)
    ```

e.  Can we say which method is a better fit to the data by comparing their log likelihood or AICc? Explain why or why not.

    *Yes, based on the results, the ARIMA model appears to be the better fit to the data when compared using AICc. The ARIMA model has a much lower AICc value than the ETS model:* $ARIMA: (AICc = -515.23726), \ ETS: (AICc = -39.18075)$

    *Since a lower AICc indicates a better balance between goodness of fit and model complexity, this suggests that the ARIMA model is preferred.*

    *The ARIMA model also has a much higher log likelihood:* $ARIMA: (log L = 266.01207), \ ETS: (log L = 38.23554)$

    *This indicates that the ARIMA model provides a substantially better in-sample fit.*

    *Therefore, for these fitted models, we would conclude that the ARIMA model is the better-fitting method. x*

## 4. Dynamic regression (adapted from Q10.4 in Hyndman & Athanasopoulos)

This exercise concerns `aus_accommodation`: the total quarterly takings from accommodation and the room occupancy level for hotels, motels, and guest houses in Australia, between January 1998 and June 2016. Total quarterly takings are in millions of Australian dollars.

```{r}
aus_accommodation

aus_accommodation |> 
  autoplot()
```

a.  Perform inflation adjustment for Takings (using the `CPI` column), creating a new column in the tsibble called `AdjTakings`.

    ```{r}
    aus_acccommodation <- aus_accommodation |>
      mutate(AdjTakings = Takings / CPI * first(CPI))

    aus_acccommodation
    ```

b.  For each state, fit a dynamic regression model of `AdjTakings` with seasonal dummy variables, a piecewise linear time trend with one knot at 2008 Q1, and ARIMA errors.

    ```{r}
    aus_accommodation <- aus_accommodation |>
      mutate(
        AdjTakings = Takings / CPI * first(CPI),
        trend = as.numeric(Date),
        knot  = pmax(as.numeric(Date) - as.numeric(yearquarter("2008 Q1")), 0)
      )

    fit <- aus_accommodation |>
      model(ARIMA(AdjTakings ~ season() + trend + knot))
    ```

c.  What model was fitted for the state of Victoria? Does the time series exhibit constant seasonality?

    ```{r}
    fit |> filter(State == "Victoria") |> report()
    ```

    *The fitted model for Victoria is a linear model with ARIMA(1,0,0)(0,0,1)\[4\] errors, incorporating seasonal dummies, a piecewise linear trend with a knot at 2008 Q1, an AR(1) term, and a seasonal MA(1) term. \
    \
    Yes, the time series exhibits constant seasonality under this model. This is because quarterly seasonality is represented using fixed seasonal dummy variables, so the seasonal effects are assumed to remain constant in magnitude over time rather than increasing or decreasing with the level of the series.*

d.  Check that the residuals of the model in c) look like white noise.

    ```{r}
    fit |> filter(State == "Victoria") |> gg_tsresiduals()

    fit |> filter(State == "Victoria") |>
      augment() |>
      features(.innov, ljung_box, lag = 8)
    ```

    *The residuals appear to behave like white noise. The residual plot shows random fluctuations around zero with no clear trend or seasonality. The ACF values are mostly within the significance bounds, indicating little remaining autocorrelation.*

    *The Ljung--Box test gave:* $[
    Q^* = 4.0099,\ p\text{-value} = 0.8562
    ]$

    *Since the p-value is greater than 0.05, we fail to reject the null hypothesis of no autocorrelation.*

    *Therefore, the residuals are consistent with white noise, suggesting that the fitted model for Victoria is adequate.*

e.  Use an ARIMA model to forecast CPI, then use these to forecast the (non-adjusted) takings for Victoria until the end of 2017. *Hint: You will have to use the `new_data` argument to `forecast()`.*

    ```{r}
    fit_cpi <- aus_accommodation |>
      filter(State == "Victoria") |>
      model(ARIMA(CPI))

    fc_cpi <- fit_cpi |> forecast(h = 6)

    new_data_vic <- new_data(
      aus_accommodation |> filter(State == "Victoria"), 6) |>
      mutate(CPI   = fc_cpi$.mean,
             trend = as.numeric(Date),
             knot  = pmax(as.numeric(Date) - as.numeric(yearquarter("2008 Q1")), 0))

    fc <- fit |>
      filter(State == "Victoria") |>
      forecast(new_data = new_data_vic) |>
      mutate(Takings_forecast = .mean * new_data_vic$CPI / 100)

    fc |> autoplot(aus_accommodation |> filter(State == "Victoria"))
    ```

f.  What sources of uncertainty have not been taken into account in the prediction intervals?

    *The prediction intervals do not account for all sources of uncertainty. They do not fully include the uncertainty in the CPI forecasts that were used as future explanatory variables, nor the uncertainty from estimating the regression and ARIMA model (knot, trend) parameters. They also ignore the possibility of future structural changes in the tourism industry, model selection uncertainty, and unexpected economic shocks or policy changes. Therefore, the actual forecast uncertainty is likely to be larger than the reported prediction intervals.*

## 5. State space models

Consider the model

$$
T_t = T_{t-1} + \epsilon_t
$$

$$
S_t = 0.8S_{t-4} + \eta_t
$$

$$
Y_t = T_t + S_t + V_t
$$

where $(\epsilon_t) \sim WN(0,\sigma_1^2)$, $\eta_t \sim WN(0,\sigma_2^2)$, and $(V_t) \sim WN(0,\sigma_3^2)$. We observe data $Y_1,\ldots,Y_n$.

a.  Write the model in the standard (matrix) form of a linear Gaussian state space model.

    $$
    X_t =
    \begin{pmatrix}
    T_t\\
    S_t\\
    S_{t-1}\\
    S_{t-2}\\
    S_{t-3}
    \end{pmatrix}
    $$

    $$
    X_t = F X_{t-1} + W_t
    $$

    $$
    F =
    \begin{pmatrix}
    1 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0.8\\
    0 & 1 & 0 & 0 & 0\\
    0 & 0 & 1 & 0 & 0\\
    0 & 0 & 0 & 1 & 0
    \end{pmatrix}
    $$

    $$
    W_t =
    \begin{pmatrix}
    \epsilon_t\\
    \eta_t\\
    0\\
    0\\
    0
    \end{pmatrix}
    $$

    $$
    Q =
    \begin{pmatrix}
    \sigma_1^2 & 0 & 0 & 0 & 0\\
    0 & \sigma_2^2 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0
    \end{pmatrix}
    $$

    $$
    Y_t = H X_t + V_t
    $$

    $$
    H =
    \begin{pmatrix}
    1 & 1 & 0 & 0 & 0
    \end{pmatrix}
    $$

    $$
    V_t \sim WN(0,\sigma_3^2)
    $$

b.  Does $\lim_{t \to \infty}\text{Var}(S_t - \hat S_{t|n})$ exist? If so, what is its value?

    $$
    S_t = 0.8S_{t-4}+\eta_t
    $$

    *Since*

    $$
    |0.8|<1,
    $$

    *the seasonal component is stationary.*

    *Let*

    $$
    \operatorname{Var}(S_t)=\gamma.
    $$

    *Then*

    $$
    \gamma = 0.8^2\gamma+\sigma_2^2
    $$

    $$
    \gamma(1-0.8^2)=\sigma_2^2
    $$

    $$
    \gamma=\frac{\sigma_2^2}{1-0.8^2}
    $$

    *Therefore,*

    $$
    \lim_{t\to\infty}\operatorname{Var}(S_t-\widehat{S}_{t|n})
    =
    \frac{\sigma_2^2}{1-0.8^2}
    $$

    $$
    =
    \frac{\sigma_2^2}{0.36}
    $$

c.  Does $\lim_{t \to \infty}\text{Var}(T_t - \hat T_{t|n})$ exist? If so, what is its value?

    $$
    T_t = T_{t-1}+\epsilon_t
    $$

    *This is a random walk, so its uncertainty increases over time.*

    *For*

    $$
    t>n,
    $$

    *we have*

    $$
    T_t = T_n+\epsilon_{n+1}+\epsilon_{n+2}+\cdots+\epsilon_t.
    $$

    *Therefore, the forecast error variance contains*

    $$
    (t-n)\sigma_1^2.
    $$

    *As*

    $$
    t\to\infty,
    $$

    *this goes to infinity.*

    *Hence,*

    $$
    \lim_{t\to\infty}\operatorname{Var}(T_t-\widehat{T}_{t|n})
    $$

    *does not exist as a finite value.*

    *Equivalently,*

    $$
    \operatorname{Var}(T_t-\widehat{T}_{t|n})\to\infty.
    $$

**AI Declaration:**

ChatGPT was used as a supplementary tool to assist with conceptual clarification and LaTeX formatting during the preparation of this assignment. All mathematical derivations, computations, and final answers were carried out by the author.
